On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2023-07-12 Manuscript/manuscript.tex
Minor edits
Went over some of Prof Schifano's notes, still need to work on Results section

Data was taken from the World Athletics 
and covers the men's 110 meter hurdles from 1999 to 2022; the source can be 
found in the Appendix.
Although women's data is available, we chose not to include women's data in the
analyses to follow due to the numerous papers citing differences in men and
women's reaction times \citep[e.g.,][]{lipps2011implications, 
babicc2009reaction, panoutsakopoulos2020gender}.
We focus on the reaction times during the semifinal and final heats only, as 
reaction times from preliminary heats are often not as fast as those in later 
heats \citep[e.g.,][]{zhang2021correlation}.

@@

\subsection{2022 National Competitions}\label{sec:datanational}
From June 23-26 2022, the United States held its Track and Field Championships 
at Hayward Field in Eugene, Oregon to decide who to send to the World Championships 
being held at the same venue in August. Thus, we can establish a baseline for the four United States athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all the World 
Track and Field Championships races compared to the United States Track and Field 
Championships races. Devon Allen raced three times at the United States Track and
Field Championships and three times at the World Track and Field Championships. 
His reaction times were substantially faster in the World Track and
Field Championships compared to the United States Track and Field Championships, 
as were his teammates. 
This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
too quickly.  Table~\ref{fig:USAvsWorld} highlights the difference in reaction 
times for Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts.

@@

The dataset of the four United States athletes 
is too small to perform any meaningful statistical analysis, but
the data was expanded to include reaction times for athletes who competed at 
national competitions (many of which were held from May-July of 2022). Once the 
data was collected, each athlete was designated as a cluster and the stage of
competition (national or international) was specified for each competition 
within the cluster (athlete).

\eds{Can you show some descriptive statistics for this data?  Total number of 
clusters? Average number of international vs national competitions across 
athletes/clusters?}
\eds{The following paragraph still needs to be smoothed (the paragraph that was 
here and the paragraph that was in the intro still need to be integrated into 
one (or two) cohesive paragraph(s) that accurately describe the two types of 
data that will be used for two rank-based analyses.}

\subsection{2019 World Track and Field Championships}
The 2019 World Track and Field Championship data is included in the larger dataset
described in section~\ref{sec:databeyond}, but we can also use it similarly to
how we used national level in section~\ref{sec:datanational}.  We can compare
the times of athletes who participated in the 2019 and 2022 World Track and Field
Championships to see if there were lower times in 2022.  While it is theoretically
possible that athletes got faster between 2019 and 2022, it is important to
remember that these are some of the fastest sprinters on the planet, and thus
improving on elite times is not easy.  Nonetheless, this is a worthy inclusion
to the data because Seiko was the timing company responsible for both events and
because the level of competition is the extremely elite in both cases.

@@

\jy{Close captions with a period.}
\begin{table}
  \centering
  \caption{Summary statistics for the GLMM on various data sets.} 
  \begin{tabular}{c c c c c c}
      \toprule
      Dataset & Venue standard error & Int effect est & Int se & Int var & Resid Var \\
      \midrule
      Old Finals & 0.0439 & $-1.901$ & 0.0251 & 0.0019 & 0.0109 \\
      All Finals & 0.0504 & $-1.914$ & 0.0280 & 0.0025 & 0.0112 \\
      Old Pooled & 0.0495 & $-1.853$ & 0.0305 & 0.0024 & 0.0243 \\
      All Pooled & 0.0554 & $-1.869$ & 0.0348 & 0.0031 & 0.0235 \\
      \bottomrule
  \end{tabular}
  \label{tab:Gamma_parameters}
\end{table}

@@

%Citation needed in this section
Since 1985 Seiko Holding Corporations has served every World Athletics 
Championships as the official timer \citep{wa2022seiko}.  Since 1985, the 
technology and the ability
to accurately predict measurements: long jump distances, false starts, total 
time, reaction time, etc. have all dramatically improved.  Seiko did not start 
tracking reaction time as an official measurement until 1999 in the men's and 
women's 100 meter \eds{110 for men?} hurdles. 
Seiko regularly updates their equipment so that they provide cutting edge
technology to the World Track and Field Championships, the highest stakes in the 
world of running outside of the Olympics. Seiko's technology for detecting 
reaction times relies on the pressure that athletes exert on the plate when they 
push off.  Their systems measure the time differential
between when the ``gun" goes off to start the race and when the pressure changes 
\citep{wa2022seiko}.  If an athlete has a reaction time under 0.1 seconds, it 
is deemed a 
false start as it is considered that no athlete can react so quickly 
\citep{Seiko-Timing}.  Thus, a reaction time of for example 0.05, suggests that 
the athlete predicted the gun and it was luck that caused their abnormally high 
time.  At the World Championships level, World Athletics and Seiko want to remove 
that element of luck and thus impose the 0.1 second barrier. In 2013, Seiko 
upgraded their timing equipment for sprinting events (110 meter hurdles falls 
under this category) \citep{wa2013backtage}. Since 1999, the
slowest reaction 
times in the men's 110-meter hurdle were recorded in 2013.  That is not to say
 that the higher 
times were caused by Seiko's equipment, but the two may be related.  It is worth 
noting that prior
to 2022, Seiko again upgraded its technology but for its jump management system.
\eds{I would argue that this paragraph should go in the discussion as there are 
no results presented here}

@@

\jy{Remind the readers what the analysis was about: two-group comparison with
  group being ....}

The rank based methods described in Section~\ref{sec:rank} are used
to determine if the reaction times for athletes who had competed at other 
championships differed significantly from the 2022 World Championships.  We can
use the 2019 World Championships and the 2022 national-level championships as a 
precedent to see how athletes performance changed across different championships.

When the clusrank rank sum test was performed, the ``ds" method produced a z score of
2.9751 and p value of 0.001464.  \eds{what is the ds method?} 
This value is for a one-sided test and show that
the mean reaction time on average in 2019 was larger (slower) than the mean 
reaction time
in 2022 for athletes who competed in both championships at an alpha level of 0.01.
This provides good evidence that athletes who competed in both 2019 and 2022 got
faster during that timespan.  \eds{That is not the only explanation}

Repeating the data collection steps and
again performing a clustered analysis for a Wilcoxon ranked sum test
results in two new z scores and p values. 
\eds{I am not sure why you are repeating the data collection steps and
again performing a clustered analysis here.  Also, if there are two new z-scores
and p values, why are you only reporting one?}
  The ``ds" method produced a z score of
3.6069 and p value of 0.000155. This result is highly significant and shows that
the times at the national stage of competition were significantly higher than
the reaction times at the World Championships even for an alpha level of 0.001.

>>>>

Data was taken from the World Athletics 
and covers the men's 110 meter hurdles from 1999 to 2022; the source can be 
found in the Appendix.
Although women's data is available, we chose not to include women's data in the
analyses to follow due to the numerous papers citing differences in men and
women's reaction times \citep[e.g.,][]{lipps2011implications, 
babicc2009reaction, panoutsakopoulos2020gender}.
For the creation of the GLMM model we focus on the reaction times during the 
semifinal and final heats only, as reaction times from preliminary heats are 
often not as fast as those in later heats \citep[e.g.,][]{zhang2021correlation}.

@@

\subsubsection{2022 National Competitions}\label{sec:datanational}
From June 23-26 2022, the United States held its Track and Field Championships 
at Hayward Field in Eugene, Oregon to decide who to send to the World Championships 
being held at the same venue in August. Thus, we can establish a baseline for the four United States athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all the World 
Track and Field Championships races compared to the United States Track and Field 
Championships races. Devon Allen raced three times at the United States Track and
Field Championships and three times at the World Track and Field Championships. 
His reaction times were substantially faster in the World Track and
Field Championships compared to the United States Track and Field Championships, 
as were his teammates. 
This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
too quickly.  Table~\ref{fig:USAvsWorld} highlights the difference in reaction 
times for Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts.

@@

The dataset of the four United States athletes 
is too small to perform any meaningful statistical analysis, but
the data was expanded to include reaction times for athletes who competed at 
national competitions (many of which were held from May-July of 2022). Once the 
data was collected, each athlete was designated as a cluster and the stage of
competition (national or international) was specified for each competition 
within the cluster (athlete).  We chose to include preliminary rounds in this
section as well as in \ref{sec:data2019} in order to increase the size of our
data.  Without data from the preliminary rounds, the number of observations
dropped from 65 to 26 and the number of athletes included in the analysis
decreased from 15 to 9.

\eds{Can you show some descriptive statistics for this data?  Total number of 
clusters? Average number of international vs national competitions across 
athletes/clusters?}
\eds{The following paragraph still needs to be smoothed (the paragraph that was 
here and the paragraph that was in the intro still need to be integrated into 
one (or two) cohesive paragraph(s) that accurately describe the two types of 
data that will be used for two rank-based analyses.}
\of{Does this work?}

\subsubsection{2019 World Track and Field Championships}\label{sec:data2019}
The 2019 World Track and Field Championship data is included in the larger dataset
described in section~\ref{sec:dataworld}, but we can also use it similarly to
how we used national level in section~\ref{sec:datanational}.  We can compare
the times of athletes who participated in the 2019 and 2022 World Track and Field
Championships to see if there were lower times in 2022.  While it is theoretically
possible that athletes got faster between 2019 and 2022, it is important to
remember that these are some of the fastest sprinters on the planet, and thus
improving on elite times is not easy.  Nonetheless, this is a worthy inclusion
to the data because Seiko was the timing company responsible for both events and
because the level of competition is the extremely elite in both cases.

@@

\jy{Close captions with a period.}
\of{Resolved}
\begin{table}
  \centering
  \caption{Summary statistics for the GLMM on various data sets.} 
  \begin{tabular}{c c c c c c}
      \toprule
      Dataset & Venue standard error & Int effect est & Int se & Int var & Resid Var \\
      \midrule
      Old Finals & 0.0439 & $-1.901$ & 0.0251 & 0.0019 & 0.0109 \\
      All Finals & 0.0504 & $-1.914$ & 0.0280 & 0.0025 & 0.0112 \\
      Old Pooled & 0.0495 & $-1.853$ & 0.0305 & 0.0024 & 0.0243 \\
      All Pooled & 0.0554 & $-1.869$ & 0.0348 & 0.0031 & 0.0235 \\
      \bottomrule
  \end{tabular}
  \label{tab:Gamma_parameters}
\end{table}

@@

%Citation needed in this section
Since 1985 Seiko Holding Corporations has served every World Athletics 
Championships as the official timer \citep{wa2022seiko}.  Since 1985, the 
technology and the ability
to accurately predict measurements: long jump distances, false starts, total 
time, reaction time, etc. have all dramatically improved.  Seiko did not start 
tracking reaction time as an official measurement until 1999 in the men's 110 meter
hurdles and women's 100 meter hurdles. 
Seiko regularly updates their equipment so that they provide cutting edge
technology to the World Track and Field Championships, the highest stakes in the 
world of running outside of the Olympics. Seiko's technology for detecting 
reaction times relies on the pressure that athletes exert on the plate when they 
push off.  Their systems measure the time differential
between when the ``gun" goes off to start the race and when the pressure changes 
\citep{wa2022seiko}.  If an athlete has a reaction time under 0.1 seconds, it 
is deemed a 
false start as it is considered that no athlete can react so quickly 
\citep{Seiko-Timing}.  Thus, a reaction time of for example 0.05, suggests that 
the athlete predicted the gun and it was luck that caused their abnormally high 
time.  At the World Championships level, World Athletics and Seiko want to remove 
that element of luck and thus impose the 0.1 second barrier. In 2013, Seiko 
upgraded their timing equipment for sprinting events (110 meter hurdles falls 
under this category) \citep{wa2013backtage}. Since 1999, the
slowest reaction 
times in the men's 110-meter hurdle were recorded in 2013.  That is not to say
 that the higher 
times were caused by Seiko's equipment, but the two may be related.  It is worth 
noting that prior
to 2022, Seiko again upgraded its technology but for its jump management system.
\eds{I would argue that this paragraph should go in the discussion as there are 
no results presented here}

@@

\jy{Remind the readers what the analysis was about: two-group comparison with
  group being ....}
\subsubsection{2019 World Track and Field Championships} \label{sec:results2019vs2022}
The rank based methods described in Section~\ref{sec:rank} are used
to determine if the reaction times for athletes who had competed at other 
championships differed significantly from the 2022 World Championships.  We can
use the 2019 World Championships and the 2022 national-level championships as a 
precedent to see how athletes performance changed across different championships.

When the clusrank rank sum test was performed, the ``ds" method produced a z score of
2.9751 and p value of 0.001464.  \eds{what is the ds method?} 
This value is for a one-sided test and show that
the mean reaction time on average in 2019 was larger (slower) than the mean 
reaction time
in 2022 for athletes who competed in both championships at an alpha level of 0.01.
This provides good evidence that athletes who competed in both 2019 and 2022 got
faster during that timespan.  \eds{That is not the only explanation}
\subsection{2022 National Championships} \label{sec:resultsnational}

  The ``ds" method produced a z score of
3.6069 and p value of 0.000155. This result is highly significant and shows that
the times at the national stage of competition were significantly higher than
the reaction times at the World Championships even for an alpha level of 0.001.
